---
title: "Prepare Dataset for Greenspace Exposure and Physical Activity"
author: "Surakshya Dhakal"
date: "2025-02-04"
output: html_document
---

<!--
NOTE:

This RMarkdown documents the exploratory and integration steps used to
construct an analysis-ready dataset combining greenspace exposure,
physical activity, and survey profile data.

It is designed for clarity and transparency, showing all intermediate steps,
checks, and summaries for readers and reviewers.

For reproducibility and automation, all key steps have been modularized into R scripts under the /R directory, which can be executed automatically via the `run_pipeline.R` script:

- `R/load_packages.R`    : Install and load all required R packages
- `R/read_data.R`        : Load raw datasets (`spe`, `grnsp`, `pa`)
- `R/clean_greenspace.R` : Compute greenspace exposure variables
- `R/prepare_pa_data.R`  : Filter, summarize, and clean physical activity data
- `R/merge_data.R`       : Merge greenspace, physical activity, and survey profile datasets
- `R/recode_variables.R` : Recategorize variables for analysis
- `R/save_data.R`        : Save final analysis-ready dataset

The fully automated pipeline is run via:
`source("run_pipeline.R")` 

This approach is taken to keep the notebook readable while separating the automated pipeline.
-->


# Summary of work done

Here’s a bullet-point summary of the critical steps taken in the script:

- **Examine Dataset Structure**:  
  - Checked the structure of the datasets (`spe`, `grnsp`, `pa`) using `str()` to understand the variables and data types.

- **Identify Unique Subjects**:  
  - Counted the unique subjects across the datasets to compare which subjects overlap in the greenspace exposure and physical activity datasets.

- **Check for Subjects with Both Greenspace and Physical Activity Data**:  
  - Identified subjects who have data in both physical activity (`pa`) and greenspace exposure (`grnsp`).

- **Check for Subjects with Only Greenspace Data**:  
  - Found subjects who have greenspace exposure data but do not have physical activity data.

- **Identify Subjects Missing in Survey Profile Dataset**:  
  - Checked for subjects who are in the greenspace and physical activity datasets but not in the survey profile dataset (`spe`).

- **Prepare Greenspace Exposure Data**:  
  - Created a new column `grn_exp` by subtracting "other_green" from "total_green".
  - Created a new column `gps_t_min` by subtracting "other_green" and "other_gray" from "total_green" and "total_gray".
  - Subset the greenspace exposure data to only include `SubjectID`, `date`, `grn_exp` and `gps_t_min`.

- **Prepare Physical Activity Data**:  
  - Subset the physical activity data to include relevant columns (`SubjectID`, `start_date`, `start_time`, etc.).
  - Reformatted date and time columns.
  - Sorted the physical activity data by `SubjectID`.

- **Handle Duplicates**:  
  - Identified and documented duplicates in the physical activity dataset, where multiple step counts are recorded for the same time intervals.
  - Clarified that these aren't true duplicates and all records should be kept.
  - Exported duplicate summaries for further inspection.

- **Filter and Clean Physical Activity Data**:  
  - Filtered physical activity data to only include time intervals between 08:00:00 and 22:00:00.
  - Summarized total steps per subject per day.
  - Summarized total wear time per subject per day.
  - Merged the summarized data with the original filtered physical activity data.

- **Merge Greenspace Exposure and Physical Activity Data**:  
  - Merged the greenspace exposure data (`grnsp_sub`) with the filtered physical activity data (`pa_sub_filtered`) on `SubjectID` and `date`.

- **Merge Survey Profile Data**:  
  - Cleaned the survey profile data (`spe`) by removing unnecessary columns and rearranging to move `age` after `SubjectID`.
  - Merged the cleaned survey profile data with the previously merged dataset.
  
- **Add Season and Weekday/Weekend Labels**:
  - Added columns for the day of the week and season (Winter, Spring, Summer, Autumn).
  - Rearranged columns for clarity.
  
- **Recategorize Employment, Income, Education, and Remote Work**:
  - Recategorized employment, income, education, and remote_work into broader categories for analysis.

- **Final Data Cleaning**:
  - Retained only necessary columns for the analysis.
  - Renamed the `start_date` column to `date` in the final dataset.
  - Sorted the data by `SubjectID` and `date` in ascending order.

- **Export Final Data**:  
  - Saved the final merged dataset to a CSV file for further analysis.


This RMarkdown file is intended for **exploratory and integration work**.  

For reproducibility and automation, all steps in this notebook have been modularized into R scripts under `/R` and can be executed sequentially using the `run_pipeline.R` script. This allows creating the same final dataset without running the notebook interactively.



# Load Packages and Set-up Working Directory

Install and load packages if required
```{r, warning=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse", "lessR", "DescTools", "kableExtra", "lubridate", "chron", "sf", "leaflet", "ggplot2", "here")
```


# Read in datasets

```{r read_data, eval=FALSE}
# Note: The raw datasets are not included in this repo.
# The code below shows how the datasets would be read and structured.

# Survey/profile dataset
spe <- read.csv("data/raw/survey_profile.csv", header = TRUE)

# Greenspace exposure dataset
grnsp <- read.csv("data/raw/greenspace_exposure.csv", header = TRUE)

# Physical activity dataset
pa <- read.csv("data/raw/physical_activity.csv", header = TRUE)
```


```{r eval=FALSE, include=FALSE}
# Examine the data structure
str(spe)
str(grnsp)
str(pa)
```


## Examine datasets
```{r}
length(unique(spe$SubjectID))     # 271 unique subjects
length(unique(grnsp$SubjectID))   # 273 unique subjects
length(unique(pa$SubjectID))      # 127 unique subjects
```


```{r}
# Subjects who have physical activity data that also have greenspace exposure data
a <- unique(pa$SubjectID)
b <- unique(pa$SubjectID) %in% unique(grnsp$SubjectID)
table(b)
sort(a[b == TRUE])
```


```{r}
# Subjects who have greenspace exposure data but not physical activity data
a <- unique(grnsp$SubjectID)
b <- unique(grnsp$SubjectID) %in% unique(pa$SubjectID)
table(b)
sort(a[b == FALSE])
```


```{r}
# Subjects not in subject profile dataset but in greenspace exposure dataset
sort(unique(grnsp$SubjectID)[unique(grnsp$SubjectID) %in% unique(spe$SubjectID) == FALSE])
```


```{r}
# Subjects not in subject profile dataset but in physical activity dataset
sort(unique(pa$SubjectID)[unique(pa$SubjectID) %in% unique(spe$SubjectID) == FALSE])
```


## Prepare greenspace data

```{r}
# Compute greenspace exposure variables
# (see R/clean_greenspace.R for details)
source(here::here("R", "clean_greenspace.R"))
grnsp <- compute_greenspace_exposure(grnsp)

# Keep only "SubjectID", "date", "grn_exp", and "gps_t"
grnsp_sub <- grnsp[, c(2, 3, 20, 21)]
```


## Prepare physical activity data

```{r}
# Subset the dataset to keep "SubjectID", "start_date", "start_time"
# "end_date", "newendtime", "newsteps", and "newdurationInSeconds"

pa_sub <- pa[, c(1:4, 19, 18, 17)]


# Reformat date and time
pa_sub <- pa_sub %>%
  mutate(start_date = as.Date(start_date, "%Y-%m-%d"),
         end_date = as.Date(end_date, "%Y-%m-%d"),
         start_time = chron(times = start_time),
         end_time = chron(times = newendtime)) %>%
  select(1:4, 8, 7, 6)   # remove "newendtime" as we are storing it with a new name


# Sort data by SubjectID
pa_sub <- pa_sub[order(pa_sub$SubjectID), ]   # Warning: data contains duplicates 
```


### Checking for duplicates
Finding: These aren't true duplicates but rather different number of steps for different activity types within the same time intervals. So, we should keep all records. 

```{r eval=FALSE}
# Check for duplicates

# Convert start_time to POSIXct for proper datetime operations
pa1 <- pa %>%
  mutate(start_datetime = ymd(start_date) + hms(start_time))

# Identify duplicates
duplicates <- pa1 %>%
  group_by(SubjectID, start_datetime) %>%
  filter(n() > 1) %>%
  ungroup()

# Remove pa1
rm(pa1)

# Count how many subjects have duplicates
num_subjects_with_duplicates <- duplicates %>%
  distinct(SubjectID) %>%
  nrow()

# Count duplicates per subject and date
duplicate_summary <- duplicates %>%
  group_by(SubjectID, start_date) %>%
  summarise(duplicate_count = n(), .groups = "drop")

# Save the duplicate summary to a CSV file
write.csv(duplicate_summary, "data/processed/duplicates_summary.csv", row.names = FALSE)

# View results
# print(num_subjects_with_duplicates)  # Number of subjects with duplicates
# print(duplicate_summary)  # Subjects and dates with duplicate records

dup_sum <- read.csv("data/processed/duplicates_summary.csv", header=TRUE)

```


```{r eval=FALSE}
# Examine a few Subjects more closely

set.seed(123)  # Set seed for reproducibility

# Get a list of unique subjects with duplicates
subjects_with_duplicates <- duplicates %>%
  distinct(SubjectID) %>%
  pull(SubjectID)

# Randomly select two subjects
random_subjects <- sample(subjects_with_duplicates, 2)

# Filter duplicates for the selected subjects
example_duplicates <- duplicates %>%
  filter(SubjectID %in% random_subjects)

# Save to CSV
write.csv(example_duplicates, "data/processed/example_duplicates.csv", row.names = FALSE)

# Print selected subjects
# print(random_subjects)

```


## Prepare physical activity data (contd.)

```{r}
# Keep time intervals between 08:00:00 and 22:00:00

# Convert start_time to a proper time format
pa_sub <- pa_sub %>%
  mutate(start_time = hms(start_time)) 

# Filter data for time between 08:00:00 and 22:00:00
pa_sub_filtered <- pa_sub %>%
  filter(start_time >= hms("08:00:00") & start_time < hms("22:00:00"))

# Summarize total steps per subject per day
daily_steps <- pa_sub_filtered %>%
  group_by(SubjectID, start_date) %>%
  summarise(total_steps = sum(newsteps, na.rm = TRUE), .groups = "drop")

# Summarize wear time per subject per day (counting each unique start_time once)
daily_wear_time <- pa_sub_filtered %>%
  group_by(SubjectID, start_date, start_time) %>%
  summarise(wear_time_per_interval = first(newdurationInSeconds), .groups = "drop") %>%
  group_by(SubjectID, start_date) %>%
  summarise(wear_t_hr = round(sum(wear_time_per_interval, na.rm = TRUE) / 3600, 2), .groups = "drop")

# Merge total steps and wear time back into "pa_sub_filtered", ensuring only one row per subject per day
pa_sub_filtered <- pa_sub_filtered %>%
  distinct(SubjectID, start_date, .keep_all = TRUE) %>%
  left_join(daily_wear_time, by = c("SubjectID", "start_date")) %>%
  left_join(daily_steps, by = c("SubjectID", "start_date"))
  

# Keep only "SubjectID", "start_date", "wear_t_hr", and "total_steps"
pa_sub_filtered <- pa_sub_filtered[, c(1, 2, 8, 9)]

```



# Merge greenspace exposure and survey profile data to physical activity

```{r}
# Merge greenspace exposure data to physical activity

# Convert "date" to Date type in both datasets
pa_sub_filtered <- pa_sub_filtered %>%
  mutate(start_date = as.Date(start_date))

grnsp_sub <- grnsp_sub %>%
  mutate(date = as.Date(date))

# Merge the two datasets based on SubjectID while matching the date
merged_df <- pa_sub_filtered %>%
  left_join(grnsp_sub, by = c("SubjectID" = "SubjectID", "start_date" = "date"))

# View the merged dataframe
# head(merged_df)
```


```{r}
# Merge survey profile data to merged_df

# Remove the columns that are not needed
spe_clean <- spe %>%
  select(-birth_date, -recruited_by, -recruited_by_code)

# Rearrange the columns to move 'age' after 'SubjectID'
spe_clean <- spe_clean %>%
  select(SubjectID, enrollment_date, age, everything())

# Rename the column 'enrollment_date' to 'enrl_date'
spe_clean <- spe_clean %>%
  dplyr::rename(enrl_date = enrollment_date)

# Reformat date and time
spe_clean <- spe_clean %>%
  mutate(enrl_date = as.Date(enrl_date, "%Y-%m-%d"))

# Merge with the previous dataframe (merged_df)
final_df <- merged_df %>%
  left_join(spe_clean, by = "SubjectID")

# View the final merged dataframe
# head(final_df)

## Rearrange columns
final_df <- final_df[, c(1, 7, 2:6, 8:22)]

```


```{r}
# Add season and weekday/weekend labels as separate columns

## Add weekday labels
final_df$wk_day <- wday(final_df$start_date, label = TRUE)   

## Rearrange columns
final_df <- final_df[, c(1:3, 23, 4:22)]

## Add season labels
final_df <- final_df %>%
  mutate(season = case_when(
    month(start_date) %in% c(12, 1, 2) ~ "w",
    month(start_date) %in% c(3, 4, 5) ~ "sp",
    month(start_date) %in% c(6, 7, 8) ~ "sum",
    month(start_date) %in% c(9, 10, 11) ~ "aut"
  ))

## Rearrange columns
final_df <- final_df[, c(1:4, 24, 5:23)]
```


```{r}
# Recategorize employment, income, education, and working from home

## Recategorize employment
final_df <- final_df %>%
  mutate(employment = case_when(
    employment %in% c("employed", "self-employed") ~ "working",
    employment %in% c("retired", "student", "other") ~ "not_working",
    TRUE ~ NA_character_  # for any unexpected cases
  ))


## Recategorize income
final_df <- final_df %>%
  mutate(income = case_when(
    income %in% c("<10K", "10-20K", "20-30K") ~ "low",
    income %in% c("30-50K", "50-75K") ~ "high",
    TRUE ~ NA_character_  # for any unexpected cases
  ))


## Recategorize education
final_df <- final_df %>%
  mutate(education = case_when(
    education %in% c("bachelor", "master", "phd") ~ "higher_ed",
    education %in% c("elementary", "secondary") ~ "lower_ed",
    TRUE ~ NA_character_  # for any unexpected cases
  ))


## Recategorize working from home
final_df <- final_df %>%
  mutate(remote_work = case_when(
    work_from_home == "0%" ~ "no",
    work_from_home %in% c("1% - 20%", "21% - 40%", "41% - 60%") ~ "low",
    work_from_home %in% c("61% - 80%", "81% - 100%") ~ "high",
    TRUE ~ NA_character_  # for any unexpected cases
  ))

```


```{r}
# Only keep following columns: "SubjectID", "enrl_date", "start_date", "wk_day", "season", ...
# .., "wear_t_hr", total_steps", "grn_exp", "gps_t_min", "age", "gender", "education", "employment", "income", and "remote_work" ...
# ... and rearrange columns for clarity.

final_df <- final_df[, c(1:11, 23, 13, 21, 25)]

```


```{r}
# Rename the column 'start_date' to 'date'
final_df <- final_df %>%
  dplyr::rename(date = start_date)

# Sort the data by 'SubjectID' and 'date' in ascending order
final_df <- final_df %>%
  arrange(SubjectID, date)

# View the updated data
# head(final_df)

```


```{r eval=FALSE}
# Export final analysis-ready dataset
# Write the final dataframe to a CSV file
write.csv(final_df, "data/processed/pa_grnsp_sp.csv", row.names = FALSE)
```

